Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-05-08 paper/main.tex
Updates from Overleaf

\note{The mapping between data and visualization is expected to be bidirectional -> that's what we really mean by equivariance and homomorphism  }
\IEEEPARstart{V}isualizations are expected to be structure preserving, meaning that when data is transformed into a graphical representation enough of the structure of the data is present in the graphic that the data can be distilled solely from the graphical representation. Changes to structure through changing the data or the graphic in the $data\leftrightarrow graphic$ mapping are expected to be reflected in each inputs corresponding output \cite{kindlmannAlgebraicProcessVisualization2014}.Because visualizations are assumed to preserve structure, visual algorithms implemented in visualization producing tools, such as software libraries, are also expected to preserve structure; frequently these expectations are either codified implicitly, only for some aspects of the data structure, or inconsistently with respect to visualization algorithm type. This means that currently many visualization tools lack an explict, comprehensive, and consistent way to express what conditions must be met to gaurantee that the component produced a structure preserving visualization.

@@

\section{Related Work}
We build on the extensive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that tools are expected to preserve when transforming data into visualizations. Structure preservation ensures that visual representations are not distortions of the data\cite{tufteVisualDisplayQuantitative2001} and it makes visualizations easier to understand \cite{norman_things_smart}. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent or compatible changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable software library components with more consistent interfaces.

@@

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. \note{add in Scales/APT/ GoG/Hullman/ETC}

\subsection{General Structure Preservation}
\note{maybe at top of section?}
Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e \in G$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.

\begin{definition}\label{def:related-work:action}\cite{nlab:action, grimaldiDiscreteCombinatorialMathematics2006}
  An \textcolor{action}{\textbf{action}} \footnote{Throughout this paper, we augment the math with color to group conceptually similar objects and functions\cite{headMathAugmentationHow2022}, for example \textcolor{action}{actions} and \textcolor{action}{sets of actions}. This color coding carries through to the figures.} of \textcolor{action}{$G$} on $X$ is a function  $act: \textcolor{action}{G} \times X \rightarrow X$. An action has the properties of identity $act(\textcolor{action}{e}, x) = x$ for all  $x \in X$ and associativity $act(\textcolor{action}{g}, act(\textcolor{action}{f}, x)) = act(\textcolor{action}{f} \circ \textcolor{action}{g}, x)$ for $\textcolor{action}{f},\textcolor{action}{g} \in \textcolor{action}{G}$.
\end{definition}

In \autoref{sec:artist:equivariant}, actions on data and compatible actions on graphics define structure in terms of testing whether a visualization is equivariant. Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$

 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}

The notion that visual encodings should be \textit{homomorphic} was proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} in his specification of \textit{A Presentation Tool} and the idea that visual transforms are \textit{equivariant} underlies Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. Kindlemann and Scheidegger propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. A uniform abstract data representation layer is introduced in \autoref{sec:atct:sheaves}, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the the correspondence between data and visual changes is included in the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist}. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. For example, connectivity assumptions tend to be embedded in each of the visual algorithms of `building block` libraries, a term used by Wongsuphasawat \cite{wongsuphasawatNavigatingWideWorld2021} to describe libraries that provide modular components for building elements of a visualization, such as functions for making boxes or translating data values to colors. In building block libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} the connectivity assumptions lead to very different interfaces; for example in Matplotlib methods for updating data and parameters for controlling aesthetics differ between plots. While VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, and therefore can embed that information in its visual algorithms, it does so in a non-uniform manner. On the other hand, domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the interface to choices that that are appropriate for the domain. For example, a tabular topological structure of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer.

@@

\begin{definition}\label{def:fiber_bundle}
   A \textbf{fiber bundle} $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and  bundle projection map $\pi: \dtotalc \rightarrow \dbasec$ \cite{FiberBundle2020,spanier1989algebraic}. 
\begin{equation}
  \label{eq:atct:fb:intro}
  \begin{tikzcd}[ampersand replacement=\&, row sep=huge]
   \dfiberc
    \arrow[r, hook, color=total] \& 
    \dtotalc
    \arrow[d, "\pi"',color=total, two heads] \\
     \& 
  \dbasec
     \arrow[u, "\dsectionc"', bend right, pos=.5, color=section, dashed]
  \end{tikzcd}
\end{equation} 
A continuous surjective map $\bm{\pi}$ is a \textbf{bundle projection} map when 
\begin{enumerate}
  \item the \textcolor{fiber}{fiber space} \dfiberc\ is the preimage of the projection function $\pi$ at a point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec\ such that $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$. All fibers in a bundle are isomorphic such that $\dfiber \cong \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$. 
  \item there is an open neighborhood $\openset_{\dbasepoint}$ surrounding each point in the base space $\dbasepoint \in \openset_{\dbasepoint} \subset \dbase$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood, denoted $\dtotal\restriction\openset$\footnote{The symbol $\restriction$ is the restriction operator\cite{RestrictionMathematics2022} defined in \texttt{amssymb}. For example $\pi^{-1}\restriction\openset$ is the inverse function $\pi^{-1}$ defined only on the points in $\openset$. Since $\pi^{-1}(\dbase)=\dtotal$, we use the shorthand $\dtotal\restriction_{\openset} \coloneqq \pi^{-1}\restriction_{\openset}$}, is locally trivial. The condition for local triviality is that $\dtotal\restriction_{\openset} = \openset \times \dfiber = \pi^{-1}\restriction_{\openset}$
\end{enumerate}
\end{definition}
\begin{definition} A \textcolor{section}{\textbf{section}} $\dsectionc: \dbasec \rightarrow \dtotalc$ over a fiber bundle is a smooth right inverse of $\pi$ such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$ for all $\dbasepoint \in \dbase$
\end{definition}

Local triviality $\dtotal\restriction_{\openset_{k}} = \openset_{k}\times \dfiber$ means that there is always a subset of the base space over which the fibers are identical. 
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{figures/local_trivalizations.png}
  \caption{The local trivializations $\dtotal_{0}$ and $\dtotal_{1}$ are subspaces of $\dtotal$ where the fibers $\dfiber_{0}$ are equal, here illustrated as all pointing in the same direction. These local trivializations are themselves locally trivial fiber bundles.} 
\end{figure}

For example, in \autoref{fig:local_trivalizations}, the bundles $\dtotal_{0}=\dtotal\restriction_{[\epsilon^{-}, {\frac{2\pi}{5}}^{+}]}$ and $\dtotal_{1}=\dtotal\restriction_{[{\frac{2\pi}{5}}^{-}, \epsilon^{-}]}$ are local trivializations of an arbitrary bundle $\dtotal$. These local trivializations overlap on the intersections of the open subsets of each trivalization $[\epsilon^{-}, \epsilon^{+}]$ and $[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]$. A set of trivializations can be glued into a larger bundle by defining transition maps that describe how to align the fibers in the overlapping space such that the sections of the bundle remain continuous. 
\note{add transition map construction formula}

@@

In a \textit{trivial bundle}, the fibers are identical $\dfiber = \dfiber_{k}$ for all points $\dbasepoint \in dbase$ such that all the transtion maps are also identical. For example, in the cylinder case the fibers in the overlappping regions of the trivalizations point in the same direction $\uparrow$ to illustrate that they are equal $\dfiber_0\restriction_{\openset_1\cap\openset_2} = \dfiber_{1}\restriction_{\openset_1\cap\openset_2}$. Because the fibers point in the same direction, the  transition maps at both intersections map the fiber to itself $\uparrow\rightarrow\uparrow$.

In a \textit{non-trivial} bundle fibers are isomorphic $\dfiber \simeq \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$ such that there is at least one transition map that is not an identity \cite{hatcherAlgebraicTopology2002}. For example, in the mobius band case, while $\dfiber_{0}\restriction_{[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]} = \dfiber_{0}\restriction_{[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]}$ such that the transition map is an identity map $\uparrow\rightarrow\uparrow$, the transition map must flip  $\uparrow \rightarrow \downarrow$ the fibers of $\dfiber_0\restriction_{[\epsilon^{-}, \epsilon^{+}]}$ in order for them to be equal to the fibers of $\dfiber_1\restriction_{[\epsilon^{-}, \epsilon^{+}]}$. 
\note{add summary sentence about power of triviality, probably pointing to it's power for abstraction}

@@

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
We encode the topological structure of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle since the base space acts as indexing space where every point in the base space has a corresponding fiber space $\pi^{-1}(\dbasepoint) = \dfiber_{\dbasepoint}$. The \textcolor{base}{base space} \dbasec\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it  consists of the set of points ${\dbasepoint \in \dbase}$ and topology $\mathcal{T}_{\dbase}$. We propose that the topology on \dbase\ can act as a mathematical abstraction for the \textcolor{base}{indexing space} of a dataset because the properties of a topology are the same as would be expected of an index space, namely that the space has a mathematical structure, that the space can be subsetted, and that continuity is preserved when sets of indices are merged. For example, in \autoref{fig:related-work:continuity:ktypes}, a topology on the line would break the line up into subsets, and those subsets could only be joined in ways where they cover the line in the same order in which it was broken up. The base space of a fiber bundle is a quotient topology\cite{QuotientSpaceTopology2020,munkresElementsAlgebraicTopology1984}, meaning that it divides the topological space into the smallest possible (largest number of) open sets such that $\pi$ remains a continuous function. This means that the topology can be defined to have a resolution equal to the number of indicies in a dataset or at whatever resolution of continuois function is required for a given task.

>>>>

\note{The mapping between data and visualization is expected to be bidirectional -> that's what we really mean by equivariance and homomorphism  }
\IEEEPARstart{V}isualizations are expected to be structure preserving, meaning that when data is transformed into a graphical representation enough of the structure of the data is present in the graphic that the data can be extracted from the graphical representation. Changes to structure through changing the data or the graphic in the $data\leftrightarrow graphic$ mapping are expected to be reflected in each inputs corresponding output \cite{kindlmannAlgebraicProcessVisualization2014}.Because visualizations are assumed to preserve structure, visual algorithms implemented in visualization producing tools, such as software libraries, are also expected to preserve structure; frequently these expectations are either codified implicitly, only for some aspects of the data structure, or inconsistently with respect to visualization algorithm type. This means that currently many visualization tools lack an explict, comprehensive, and consistent way to express what conditions must be met to gaurantee that the component produced a structure preserving visualization.

@@

\section{Related Work}
We build on extensive work codifying the structure of visualization data and corresponding graphics and use the concepts of \textit{homomorphism} and \textit{equivariance} to express how that structure should be preserved. Structure is considered preserved when the input and output change in tandem, and these symmetries are often modeled as \textit{group actions}, which are a way for a group\footnote{A \textit{group} is a set with an associative binary operator, an identity, and an inverse operation} to transform some other set:

\begin{definition}\label{def:related-work:action}\cite{nlab:action, grimaldiDiscreteCombinatorialMathematics2006}
  An \textcolor{action}{\textbf{action}} \footnote{Throughout this paper, we augment the math with color to group conceptually similar objects and functions\cite{headMathAugmentationHow2022}, for example \textcolor{action}{actions} and \textcolor{action}{sets of actions}. This color coding carries through to the figures.} of \textcolor{action}{$G = (G,\circ, e)$} on $X$ is a function  $act: \textcolor{action}{G} \times X \rightarrow X$. An action has the properties of identity $act(\textcolor{action}{e}, x) = x$ for all  $x \in X$ and associativity $act(\textcolor{action}{g}, act(\textcolor{action}{f}, x)) = act(\textcolor{action}{f} \circ \textcolor{action}{g}, x)$ for $\textcolor{action}{f},\textcolor{action}{g} \in \textcolor{action}{G}$.
\end{definition}

Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$

 \begin{definition}
$f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}

 \note{add something back in for homomorphic}

The notion that visual encodings should be \textit{homomorphic} was proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} in his specification of \textit{A Presentation Tool} and the idea that visual transforms are \textit{equivariant} underlies Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. Kindlemann and Scheidegger propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. A uniform abstract data representation layer is introduced in \autoref{sec:atct:sheaves}, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the the correspondence between data and visual changes is included in the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist}. We define structure preservation both in \note{field types and topology}

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
 Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. \note{add in Scales/APT/ GoG/Hullman/ETC}

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. For example, connectivity assumptions tend to be embedded in each of the visual algorithms of `building block` libraries, a term used by Wongsuphasawat \cite{wongsuphasawatNavigatingWideWorld2021} to describe libraries that provide modular components for building elements of a visualization, such as functions for making boxes or translating data values to colors. In building block libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} the connectivity assumptions lead to very different interfaces; for example in Matplotlib methods for updating data and parameters for controlling aesthetics differ between plots. While VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, and therefore can embed that information in its visual algorithms, it does so in a non-uniform manner. On the other hand, domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the interface to choices that that are appropriate for the domain. For example, a tabular topological structure of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer. 
\note{transition sentence}
In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

The fiber bundle abstraction is a structure fusing some \textcolor{base}{base space} (indexing into the data) with some \textcolor{fiber}{fiber space} (acting as the data domain). This sort of fusion could be done by a simple direct product (or join-operation), but with the fiber bundle abstraction we get the flexibility of expressing more complicated fusions than the direct product, as long as they can be constructed from building blocks that look like direct products.

\begin{definition}\label{def:fiber_bundle}
   A \textbf{fiber bundle} $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and  bundle projection map $\pi: \dtotalc \rightarrow \dbasec$ \cite{FiberBundle2020,spanier1989algebraic}. 
\begin{equation}
  \label{eq:atct:fb:intro}
  \begin{tikzcd}[ampersand replacement=\&, row sep=huge]
   \dfiberc
    \arrow[r, hook, color=total] \& 
    \dtotalc
    \arrow[d, "\pi"',color=total, two heads] \\
     \& 
  \dbasec
     \arrow[u, "\dsectionc"', bend right, pos=.5, color=section, dashed]
  \end{tikzcd}
\end{equation} 
A continuous surjective map $\bm{\pi}$ is a \textbf{bundle projection} map when 
\begin{enumerate}
  \item all fibers in the bundle are isomorphic. Since all fibers are isomorphic, there is a uniquely determined \textcolor{fiber}{fiber space} \dfiberc\ given by the preimage of the projection $\pi$ at any point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec: $\dfiberc = \pi^{-1}(k)$.  
  %the \textcolor{fiber}{fiber space} \dfiberc\ is the preimage of the projection function $\pi$ at a point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec\ such that $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$. All fibers in a bundle are isomorphic such that $\dfiber \cong \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$. 
  \item each point $\dbasepointc$ in the \textcolor{base}{base space} \dbasec\ has an open neighborhood $\openset_{\dbasepointc}$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood is locally trivial. \textbf{Local triviality} means $\dtotal\restriction_{\openset} = \openset\times F$. In this paper we use $\dtotal\restriction_{\openset} = \pi^{-1}(\openset)$ to denote the preimage of an openset, and a \textbf{local trivialization} is a specific choice of neighborhoods and their preimages.
  %there is an open neighborhood $\openset_{\dbasepoint}$ surrounding each point in the base space $\dbasepoint \in \openset_{\dbasepoint} \subset \dbase$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood, denoted $\dtotal\restriction\openset$\footnote{The symbol $\restriction$ is the restriction operator\cite{RestrictionMathematics2022} defined in \texttt{amssymb}. For example $\pi^{-1}\restriction\openset$ is the inverse function $\pi^{-1}$ defined only on the points in $\openset$. Since $\pi^{-1}(\dbase)=\dtotal$, we use the shorthand $\dtotal\restriction_{\openset} \coloneqq \pi^{-1}\restriction_{\openset}$}, is locally trivial. The condition for local triviality is that $\dtotal\restriction_{\openset} = \openset \times \dfiber = \pi^{-1}\restriction_{\openset}$
\end{enumerate}
\end{definition}
\begin{definition} A \textcolor{section}{\textbf{section}} $\dsectionc: \dbasec \rightarrow \dtotalc$ over a fiber bundle is a smooth right inverse of $\pi$: $\pi(\dsection(\dbasepoint)) = \dbasepoint$ for all $\dbasepoint \in \dbase$
\end{definition}

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{figures/local_trivalizations.png}
  \caption{The local trivializations $\dtotal_{0}$ and $\dtotal_{1}$ are subspaces of $\dtotal$ where the fibers $\dfiber_{0}$ are equal, here illustrated as all pointing in the same direction. These local trivializations are themselves locally trivial fiber bundles.} 
  \label{fig:local_trivalizations}
\end{figure}

\note{Add motiviating note here!}
Consider the circle. Points can be identified with angles (in radians, for instance), and we can imagine covering the circle with two overlapping intervals - for instance $\dbase_0=(-\varepsilon, 2\pi/5+\varepsilon)$ and $\dbase_1=(2\pi/5-\varepsilon, 2\pi+\varepsilon)$. We can imagine creating building blocks $\dbase_0\times[-1,1]$ and $\dbase_1\times[-2,2]$.

There are two different ways to create a fiber bundle that has these building blocks as the local trivializations: on the one hand, we could imagine just a cylinder $S^1\times[-1,1]$ as a total space. The fibers would all be just the interval $[-1,1]$. Over $\dbase_0$, we could map from the local trivialization to the total space with the identity map, and over $\dbase_1$ we could map from the local trivialization to the total space by rescaling the fiber space $[-2,2]\to[-1,1]$.

On the other hand, we could imagine creating a Möbius strip by flipping one end of $\dbase_1\times[-2,2]$ over before attaching it, so that while $\dbase_0\times[-1,1]$ gets mapped into the total space with the identity map, the part $(2\pi-\varepsilon,2\pi+\varepsilon)\times[-2,2]$ of $\dbase_1\times[-2,2]$ is mapped by the function $x\mapsto -x/2$ instead of the function $x\mapsto x/2$.

These functions - using the isomorphisms $x\mapsto x$, $x\mapsto x/2$ and $x\mapsto -x/2$ to map from each rectangle into the cylinder compose into different maps $(\dbase_0\cap\dbase_1)\times[-2,2]\to(\dbase_0\cap\dbase_1)\times[-1,1]$ where the source is seen as a subset of $\dbase_1\times[-2,2]$ and the target is seen as a subset of $\dbase_0\times[-1,1]$. By specifying just these \textbf{transition maps} for all overlaps of cover elements in the local trivialization we can specify a gluing scheme that constructs the fiber bundle from these locally trivial pieces.

%For example, in \autoref{fig:local_trivalizations}, the bundles $\dtotal_{0}=\dtotal\restriction_{[\epsilon^{-}, {\frac{2\pi}{5}}^{+}]}$ and $\dtotal_{1}=\dtotal\restriction_{[{\frac{2\pi}{5}}^{-}, \epsilon^{-}]}$ are local trivializations of an arbitrary bundle $\dtotal$. These local trivializations overlap on the intersections of the open subsets of each trivalization $[\epsilon^{-}, \epsilon^{+}]$ and $[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]$. A set of trivializations can be glued into a larger bundle by defining transition maps that describe how to align the fibers in the overlapping space such that the sections of the bundle remain continuous. 
\note{add transition map construction formula}

@@

A \textit{trivial bundle} is directly isomorphic to $\dbasec\times\dfiberc$, and for any choice of cover of $\dbasec$ by overlapping opensets, we can choose local trivializations so that all transition maps are identity maps. In the example in Figure~\ref{fig:cyl_mob_bundles}, we use arrows $\uparrow$ to denote fiber alignments, illustrating that by choosing simple enough trivializations, we get identity maps in the transition maps.

%In a \textit{trivial bundle}, the fibers are identical $\dfiber = \dfiber_{k}$ for all points $\dbasepoint \in dbase$ such that all the transtion maps are also identical. For example, in the cylinder case the fibers in the overlappping regions of the trivalizations point in the same direction $\uparrow$ to illustrate that they are equal $\dfiber_0\restriction_{\openset_1\cap\openset_2} = \dfiber_{1}\restriction_{\openset_1\cap\openset_2}$. Because the fibers point in the same direction, the  transition maps at both intersections map the fiber to itself $\uparrow\rightarrow\uparrow$.

A \textit{non-trivial bundle} on the other hand is one that can not be constructed as $\dbasec\times\dfiberc$, but where the construction is more complicated. Then for any choice of local trivializations, there is at least one transition map that is not an identity \cite{hatcherAlgebraicTopology2002}. For example, in the Möbius band case, while $\dfiber_0\restriction_{(2\pi/5-\varepsilon, 2\pi/5+\varepsilon)}\to\dfiber_1\restriction_{(2\pi/5-\varepsilon, 2\pi/5+\varepsilon)}$ can be chosen to be an identity map (after transforming $\dbase_1\times[-2,2]\to\dbase_1\times[-1,1]$ by rescaling), the other transition map component $\dfiber_0\restriction_{(-\varepsilon,\varepsilon)}\to\dfiber_1\restriction_{(-\varepsilon,\varepsilon)}$ has to flip the fibers to align them between the pieces.

%In a \textit{non-trivial} bundle fibers are isomorphic $\dfiber \simeq \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$ such that there is at least one transition map that is not an identity \cite{hatcherAlgebraicTopology2002}. For example, in the mobius band case, while $\dfiber_{0}\restriction_{[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]} = \dfiber_{0}\restriction_{[{\frac{2\pi}{5}}^{-}, {\frac{2\pi}{5}}^{+}]}$ such that the transition map is an identity map $\uparrow\rightarrow\uparrow$, the transition map must flip  $\uparrow \rightarrow \downarrow$ the fibers of $\dfiber_0\restriction_{[\epsilon^{-}, \epsilon^{+}]}$ in order for them to be equal to the fibers of $\dfiber_1\restriction_{[\epsilon^{-}, \epsilon^{+}]}$.

@@

\subsubsection{Topological Structure: Base Space \dbase}
\label{sec:atct:fb:base}
\note{base space of bundle for key/value pair, topology in general 'cause data + viz continuity in same way}
Munzner~\cite{munznerChDataAbstraction} describes a key-value semantics for data representation and abstraction, where datasets can be queried with different kinds of keys depending on the dataset type - locations for values of spatial fields, times for values of timeseries, or indexes for values in flat or multidimensional tables.
We propose that these different types of keys are instances of different underlying topological organization of the data, and that we can use the structure of the opensets in the topological space to encode the extent to which the keys can vary continuously: flat or multidimensional tables have discrete points as indices, timeseries have points on a line as indices and spatial fields (regardless of whether they are scalar or tensor fields) have points in a spatial domain as indices. Therefore we choose to encode the topological structure of the data as the \textcolor{base}{base space} of a fiber bundle. Since every point in the base space has a corresponding fiber space that contains the data values, we preserve the general notion of a key (continuity)-value (data) pairing. We further follow the example set by \cite{spivakSimplicialDatabases2009,spivakDatabasesAreCategories2010} and propose that the structure of the data types be formally specified as the objects of a category.

%We encode the topological structure of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle since the base space acts as indexing space where every point in the base space has a corresponding fiber space $\pi^{-1}(\dbasepoint) = \dfiber_{\dbasepoint}$. The \textcolor{base}{base space} \dbasec\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it  consists of the set of points ${\dbasepoint \in \dbase}$ and topology $\mathcal{T}_{\dbase}$. We propose that the topology on \dbase\ can act as a mathematical abstraction for the \textcolor{base}{indexing space} of a dataset because the properties of a topology are the same as would be expected of an index space, namely that the space has a mathematical structure, that the space can be subsetted, and that continuity is preserved when sets of indices are merged. For example, in \autoref{fig:related-work:continuity:ktypes}, a topology on the line would break the line up into subsets, and those subsets could only be joined in ways where they cover the line in the same order in which it was broken up. The base space of a fiber bundle is a quotient topology\cite{QuotientSpaceTopology2020,munkresElementsAlgebraicTopology1984}, meaning that it divides the topological space into the smallest possible (largest number of) open sets such that $\pi$ remains a continuous function. This means that the topology can be defined to have a resolution equal to the number of indicies in a dataset or at whatever resolution of continuois function is required for a given task.

%The topological structure of the data can be expressed programmatically by constructing data types that encapsulate the class of topological structure-.i.e point, line, plane, network, etc. We propose that these data types can be formally specified as objects of a category.
